---
title: "DSPA_LiP_and_TC"
output:
  html_document: default
  pdf_document: default
  word_document: default
date: "2024-09-18"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Preperation

## Loading packages 

```{r loading_packages, message=FALSE, warning=FALSE}
# Installing the required packages
if (!requireNamespace("dendextend", quietly = TRUE)) install.packages("dendextend")
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
if (!requireNamespace("seriation", quietly = TRUE)) install.packages("seriation")

# Loading the libraries
library(dendextend)
library(pheatmap)
library(seriation)
library(protti)
library(tidyverse)
library(data.table)
library(dplyr)
library(magrittr)
library(stringr)
library(tidyr)
library(ggplot2)
library(gridExtra)


```

## Loading data


```{r loading_data, message=FALSE, warning=FALSE}

setwd("/Volumes/biol_bc_picotti_1/Maximilian_D/026_AT9084_Re/SN")

DIA_raw <- read_protti("20240712_093839_mduennebacke_Report_MD.csv")
```
## filter data

```{r cleaning_data, message = FALSE, warning = FALSE}


DIA_raw$intensity_log2 <- log2(DIA_raw$fg_ms2raw_quantity)


DIA_clean <- DIA_raw %>%
  #filter(pg_fasta_files != "contaminants", pg_fasta_files != "NaN") %>%
  filter(intensity_log2 > 10) %>%
  #mutate(intensity_log2 = log2(fg_quantity)) %>%
  filter(pep_is_proteotypic == T)# %>%
  #mutate(uniprot_id = pg_protein_accessions)





DIA_clean_norm <- normalise(DIA_clean,
  sample = r_file_name,
  intensity_log2 = intensity_log2,
  method = "median"
)



DIA_clean_norm$fg_id <- paste0(DIA_clean_norm$fg_labeled_sequence, DIA_clean_norm$fg_charge)



unis <- unique(DIA_clean_norm$pg_protein_accessions) # make vector for fetch_uniprot
```



## Load data from uniprot and join with DIA dataframe

```{r uniprot, message = FALSE, warning = FALSE}

uniprot <- # download protein information from UniProt
  fetch_uniprot(
    unis,
    columns =  c(
      "protein_name",
      "gene_names",
      "length",
      "sequence",
      "xref_pdb"
    )
  ) #%>%
  # rename(
  #   protein_sequence = sequence,
  #   length_protein = length,
  #   uniprot_id = accession
  # )

DIA_clean_uniprot <- DIA_clean_norm %>%
  left_join(uniprot, by = c("pg_protein_accessions" = "accession")) %>% # rejoin with annotations
  find_peptide(sequence, pep_stripped_sequence) %>%
  assign_peptide_type(aa_before, last_aa, aa_after) %>%
  # sequence_coverage(sequence, pep_stripped_sequence) %>%
  distinct()

DIA_clean_uniprot$condrep <- paste(DIA_clean_uniprot$r_condition, DIA_clean_uniprot$r_replicate, sep = "_")

proteins_identified <- uniprot %>%
  distinct(accession)

# fwrite(proteins_identified, file = "identified_proteins_CaM_LiP.csv", sep = ",")


```
##Imputation on precursor lvl 


```{r imputation_on_precursor_lvl, message = FALSE, warning = FALSE}



#needs to be added 


```


##sum up precursors to peptide lvl and keep only one entry per pep_stripped_sequence
###code here needs to be adjusted for protti function

```{r sum_and_Filter_for_stripped_sequence, message = FALSE, warning = FALSE}




DIA_clean_uniprot_summed <- DIA_clean_uniprot %>%
  mutate(intensity = 2^normalised_intensity_log2) %>%
  group_by(condrep, pep_stripped_sequence) %>%
  mutate(summed_intensity = sum(intensity), .keep_all = TRUE) %>%
  mutate(summed_intensity_log2 = log2(summed_intensity)) %>%
  distinct(pep_stripped_sequence, .keep_all = TRUE) %>%
  ungroup()




DIA_clean_uniprot_summed <- DIA_clean_uniprot_summed %>%
  group_by(r_file_name, pep_stripped_sequence) %>%
  slice(1) %>%
  ungroup() %>%
  select(-starts_with("fg_"), -starts_with("eg_"))

```













#QC

##CV

```{r quality_control_cvs, fig.width = 12, fig.align= "center",  message = FALSE, warning = FALSE}

qc_cvs(
  data = DIA_clean_uniprot,
  grouping = fg_id,
  condition = r_condition,
  intensity = fg_quantity,
  plot = FALSE
)

qc_cvs(
  data = DIA_clean_uniprot,
  grouping = fg_id,
  condition = r_condition,
  intensity = fg_quantity,
  plot_style = "density",
  plot = TRUE
)

qc_cvs(
  data = DIA_clean_uniprot,
  grouping = fg_id,
  condition = r_condition,
  intensity = fg_quantity,
  plot_style = "violin",
  plot = TRUE
)

```

## Intensity distribution

Intensity distributions are plotted for the whole dataset.

```{r quality_control_log2_int_distr, fig.align= "center", message = FALSE, warning = FALSE}


qc_intensity_distribution(
  DIA_clean_uniprot,
  condrep,
  pep_grouping_key,
  intensity_log2,
  plot_style = "histogram"
)

```

## Missed cleavages


```{r quality_control_missed_cleav, fig.align= "center", message = FALSE, warning = FALSE}
qc_missed_cleavages(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_nr_of_missed_cleavages,
  fg_quantity,
  method = "intensity",
  plot = TRUE,
  interactive = FALSE
)

qc_missed_cleavages(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_nr_of_missed_cleavages,
  fg_quantity,
  method = "count",
  plot = TRUE,
  interactive = FALSE
)
```

## Peptide types

Peptide type (tryptic, semi-tryptic, non-tryptic) distributions are even throughout the different samples.

```{r quality_control_pep_type, fig.align = "center", message = FALSE, warning = FALSE}

qc_peptide_type(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_type,
  intensity = fg_quantity,
  method = "count",
  plot = TRUE,
  interactive = FALSE
)

qc_peptide_type(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_type,
  intensity = fg_quantity,
  method = "intensity",
  plot = TRUE,
  interactive = FALSE
)

```

## Number of peptide IDs per sample

The numbers of identified peptides are consistent throughout the different samples.

```{r quality_control_pep_ID, fig.align = "center", message = FALSE, warning = FALSE}

DIA_raw$condrep <- paste(DIA_raw$r_condition, DIA_raw$r_replicate, sep = "_")


qc_ids(DIA_raw, condrep, pep_grouping_key, condition = r_condition, intensity = fg_quantity)

```

## Principal component analysis (PCA)

```{r PCA, message = FALSE, warning = FALSE}

DIA_clean_uniprot %>%
qc_pca(condrep, 
       pep_grouping_key, 
       intensity_log2, 
       r_condition)

```



## corelation_map


```{r correlation_map, message = FALSE, warning = FALSE}

qc_sample_correlation(
  data = DIA_clean_uniprot,
  sample = r_file_name,
  grouping = fg_id,
  intensity_log2 = intensity_log2,
  condition = r_condition
)

```







#Data analysis
##LiP
### Volcano plots peptide lvl


```{r Volcano_plot, message = FALSE, warning = FALSE}

  


Volcano_input <- DIA_clean_uniprot_summed %>%
  unique() %>%
  assign_missingness(r_file_name,
                     r_condition,
                     pep_stripped_sequence,
                     normalised_intensity_log2,
                     ref_condition = "0_mM",
                     retain_columns = c(pg_protein_accessions, pep_stripped_sequence, start, end, pep_type, sequence, length))


Volcano_input = Volcano_input %>%
  dplyr::filter(is.na(sequence)== F)



#differential analysis

t_test_result_mod_t_test_pep <- calculate_diff_abundance(data = Volcano_input,
                                r_file_name,
                                r_condition,
                                pep_stripped_sequence,
                                normalised_intensity_log2,
                                missingness,
                                comparison = comparison,
                                ref_condition = "0_mM",
                                method = "moderated_t-test",
                                retain_columns = c(pg_protein_accessions, pep_stripped_sequence, start, end, pep_type, sequence, length))



condition_vulcano <- c("10_mM_vs_0_mM")
t_test_result_mod_t_test_pep_comp <- t_test_result_mod_t_test_pep[t_test_result_mod_t_test_pep$comparison %in% condition_vulcano, ]


volcano_plot(t_test_result_proDA_pep_comp, pep_stripped_sequence, diff, adj_pval, method = "target", target_column = pg_protein_accessions, target = c("P26365") ,  x_axis_label = "log2(fold change) Mercapto vs. DMSO", title = "Volcano plot peptide level", y_axis_label = "-log10(q-value)", log2FC_cutoff = 1, significance_cutoff = 0.05, interactive = T)



# Specify the output path
output_path_diff_abundance <- "add/here.csv" 

# Save the dataframe as a CSV file
write.csv(t_test_result_mod_t_test_pep, file = output_path_diff_abundance, row.names = FALSE)


```


##TC



###preparation


####loading data

```{r loading_data_TC, message=FALSE, warning=FALSE}

setwd("set/to/target/location")

DIA_raw_TC <- read_protti("filename.csv")
```


####cleaning
```{r cleaning_data_TC, message = FALSE, warning = FALSE}


#Do we use Spectronaut protein quantity here?????? otherwise, we need to adjust here
DIA_raw_TC$intensity_log2 <- log2(DIA_raw_TC$pg_quantity)
DIA_raw_TC$precursor_intensity_log2 <- log2(DIA_raw_TC$fg_ms2raw_quantity)




DIA_clean_TC <- DIA_raw_TC %>%
  #filter(pg_fasta_files != "contaminants", pg_fasta_files != "NaN") %>%
  filter(precursor_intensity_log2 > 10) %>%
  #mutate(intensity_log2 = log2(fg_quantity)) %>%
  filter(pep_is_proteotypic == T)# %>%
  #mutate(uniprot_id = pg_protein_accessions)






DIA_clean_norm_TC <- normalise(DIA_clean_TC,
  sample = r_file_name,
  intensity_log2 = intensity_log2,
  method = "median"
)





unis_TC <- unique(DIA_clean_TC$pg_protein_accessions) # make vector for fetch_uniprot
```


####uniprot

```{r uniprot_TC, message = FALSE, warning = FALSE}

uniprot_TC <- # download protein information from UniProt
  fetch_uniprot(
    unis_TC,
    columns =  c(
      "protein_name",
      "gene_names",
      "length",
      "sequence",
      "xref_pdb",
      "go_f"
    )
  ) #%>%
  # rename(
  #   protein_sequence = sequence,
  #   length_protein = length,
  #   uniprot_id = accession
  # )

DIA_clean_uniprot_TC <- DIA_clean_norm_TC %>%
  left_join(uniprot_TC, by = c("pg_protein_accessions" = "accession")) %>% # rejoin with annotations
  find_peptide(sequence, pep_stripped_sequence) %>%
  assign_peptide_type(aa_before, last_aa, aa_after) %>%
  # sequence_coverage(sequence, pep_stripped_sequence) %>%
  distinct()

DIA_clean_uniprot_TC$condrep <- paste(DIA_clean_uniprot_TC$r_condition, DIA_clean_uniprot_TC$r_replicate, sep = "_")

proteins_identified_TC <- uniprot_TC %>%
  distinct(accession)

# fwrite(proteins_identified, file = "identified_proteins_CaM_LiP.csv", sep = ",")


```





####volcano plot



```{r Volcano_plot_TC, message = FALSE, warning = FALSE}

#filter to only keep one entry per protein (if we don't use spectronaut quantities, we might want to change this one as well)
DIA_clean_uniprot_pep_filtered_TC <- DIA_clean_uniprot_TC %>%
  group_by(condrep) %>%
  distinct(pg_protein_accessions, .keep_all = TRUE) %>%
  ungroup() %>% # Optional, to remove the grouping structure if no longer needed
  select(-starts_with("fg_"), -starts_with("eg_"), -starts_with("pep_"))  # Drop columns starting with fg_ or eg_


DIA_clean_uniprot_pep_filtered_TC %>%
  filter(!is.na(pg_protein_accessions) & pg_protein_accessions != "") %>%
  summarise(count = n())


Volcano_input_TC <- DIA_clean_uniprot_pep_filtered_TC %>%
  unique() %>%
  assign_missingness(r_file_name,
                     r_condition,
                     pg_protein_accessions,
                     normalised_intensity_log2,
                     ref_condition = "DMSO_TC",
                     retain_columns = c(pg_protein_accessions, go_f))


t_test_result_t_test_pep_TC <- calculate_diff_abundance(data = Volcano_input_TC,
                                r_file_name,
                                r_condition,
                                pg_protein_accessions,
                                normalised_intensity_log2,
                                missingness,
                                comparison = comparison,
                                ref_condition = "DMSO_TC",
                                method = "t-test",
                                retain_columns = c(pg_protein_accessions, go_f))





condition_vulcano <- c("Aza_TC_vs_DMSO_TC")
t_test_result_t_test_pep_comp_TC <- t_test_result_t_test_pep_TC[t_test_result_t_test_pep_TC$comparison %in% condition_vulcano, ]


volcano_plot(t_test_result_t_test_pep_comp_TC, pep_grouping_key, diff, pval, method = "target", target_column = pg_protein_accessions, target = c("P0AF18", "P0A759") ,  x_axis_label = "log2(fold change) ", title = "Volcano plot precursor level", y_axis_label = "-log10(p-value)", log2FC_cutoff = 1, significance_cutoff = 0.01, interactive = T)




```


####safe output

```{r save_to_CSV_TC, fig.align = "center", message = FALSE, warning = FALSE}



# Specify the output path
output_path_diff_abundance_TC <- "outpu/path/and/name.csv" 

# Save the dataframe as a CSV file
write.csv(t_test_result_t_test_pep_TC, file = output_path_diff_abundance_TC, row.names = FALSE)


```



###data analysis

####Protein lvl adjustment


```{r diff_prot_adjust, fig.align = "center", message = FALSE, warning = FALSE}





t_test_result_t_test_pep_TC <- t_test_result_t_test_pep_TC %>% 
  mutate(comparison = str_replace_all(comparison, "_TC", ""))

# t_test_result_mod_t_test_pep <- t_test_result_mod_t_test_pep %>% 
  # mutate(comparison = str_replace_all(comparison, "_LiP", ""))

corrected_pg <- correct_lip_for_abundance(
  lip_data = t_test_result_mod_t_test_pep,
  trp_data = t_test_result_t_test_pep_TC,
  protein_id = pg_protein_accessions,
  grouping = pep_stripped_sequence,
  retain_columns = c("missingness"),
  method = "satterthwaite"
)





condition_vulcano_adj <- c("Aza_vs_DMSO")
corrected_pg_comp <- corrected_pg[corrected_pg$comparison %in% condition_vulcano_adj, ]



volcano_plot(corrected_pg_comp, pg_protein_accessions, diff, adj_pval, method = "target", target_column = pg_protein_accessions, target = c("P77258") ,  x_axis_label = "log2(fold change) ", title = "Volcano plot peptide level", y_axis_label = "-log10(p-value)", log2FC_cutoff = 1, significance_cutoff = 0.01, interactive = T)



```









##save output

```{r save_to_CSV_adj, fig.align = "center", message = FALSE, warning = FALSE}



# Specify the output path
output_path_adj <- "outpu/path/and/name.csv" 

# Save the dataframe as a CSV file
write.csv(corrected_pg, file = output_path_adj, row.names = FALSE)


```




