---
title: "DSPA_LiP_Quant"
output:
  html_document: default
  pdf_document: default
  word_document: default
date: "2024-09-18"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Preperation

## Loading packages 

```{r loading_packages, message=FALSE, warning=FALSE}
# Installing the required packages
if (!requireNamespace("dendextend", quietly = TRUE)) install.packages("dendextend")
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
if (!requireNamespace("seriation", quietly = TRUE)) install.packages("seriation")

# Loading the libraries
library(dendextend)
library(pheatmap)
library(seriation)
library(protti)
library(tidyverse)
library(data.table)
library(dplyr)
library(magrittr)
library(stringr)
library(tidyr)
library(ggplot2)
library(gridExtra)


```

## Loading data


```{r loading_data, message=FALSE, warning=FALSE}

DIA_raw <- read_protti("./data/mduennebacke_Report_MD_20240917_135800.tsv")
```
## filter data

```{r cleaning_data, message = FALSE, warning = FALSE}


# Either filter out peptides during potential ion suppression 

DIA_raw <- DIA_raw[!(DIA_raw$eg_apex_rt > 0 & DIA_raw$eg_apex_rt < 35), ]

# OR Filter the data frame to exclude entire conc. with ion suppression


#DIA_raw$concentration_mM <- as.numeric(sub("_mM", "", DIA_raw$r_condition))
#DIA_raw <- DIA_raw[DIA_raw$concentration_mM <= 0.05, ]



#hist(DIA_raw$eg_apex_rt)

DIA_raw$intensity_log2 <- log2(DIA_raw$fg_ms2raw_quantity)


DIA_clean <- DIA_raw %>%
  #filter(pg_fasta_files != "contaminants", pg_fasta_files != "NaN") %>%
  filter(intensity_log2 > 10) %>%
  #mutate(intensity_log2 = log2(fg_quantity)) %>%
  filter(pep_is_proteotypic == T)# %>%
  #mutate(uniprot_id = pg_protein_accessions)





DIA_clean_norm <- normalise(DIA_clean,
  sample = r_file_name,
  intensity_log2 = intensity_log2,
  method = "median"
)



DIA_clean_norm$fg_id <- paste0(DIA_clean_norm$fg_labeled_sequence, DIA_clean_norm$fg_charge)






unis <- unique(DIA_clean_norm$pg_protein_accessions) # make vector for fetch_uniprot
```



## Load data from uniprot and join with DIA dataframe

```{r uniprot, message = FALSE, warning = FALSE}

uniprot <- # download protein information from UniProt
  fetch_uniprot(
    unis,
    columns =  c(
      "protein_name",
      "gene_names",
      "length",
      "sequence",
      "xref_pdb"
    )
  ) #%>%
  # rename(
  #   protein_sequence = sequence,
  #   length_protein = length,
  #   uniprot_id = accession
  # )

DIA_clean_uniprot <- DIA_clean_norm %>%
  left_join(uniprot, by = c("pg_protein_accessions" = "accession")) %>% # rejoin with annotations
  find_peptide(sequence, pep_stripped_sequence) %>%
  assign_peptide_type(aa_before, last_aa, aa_after) %>%
  # sequence_coverage(sequence, pep_stripped_sequence) %>%
  distinct()

DIA_clean_uniprot$condrep <- paste(DIA_clean_uniprot$r_condition, DIA_clean_uniprot$r_replicate, sep = "_")

proteins_identified <- uniprot %>%
  distinct(accession)

# fwrite(proteins_identified, file = "identified_proteins_CaM_LiP.csv", sep = ",")


```




##sum up precursors to peptide lvl and keep only one entry per pep_stripped_sequence
###code here needs to be adjusted for protti function

```{r sum_and_Filter_for_stripped_sequence, message = FALSE, warning = FALSE}




DIA_clean_uniprot_summed <- DIA_clean_uniprot %>%
  mutate(intensity = 2^normalised_intensity_log2) %>%
  group_by(condrep, pep_stripped_sequence) %>%
  mutate(summed_intensity = sum(intensity), .keep_all = TRUE) %>%
  mutate(summed_intensity_log2 = log2(summed_intensity)) %>%
  distinct(pep_stripped_sequence, .keep_all = TRUE) %>%
  ungroup()




DIA_clean_uniprot_summed <- DIA_clean_uniprot_summed %>%
  group_by(r_file_name, pep_stripped_sequence) %>%
  slice(1) %>%
  ungroup() %>%
  select(-starts_with("fg_"), -starts_with("eg_"))

```













#QC

##CV

```{r quality_control_cvs, fig.width = 12, fig.align= "center",  message = FALSE, warning = FALSE}

qc_cvs(
  data = DIA_clean_uniprot,
  grouping = fg_id,
  condition = r_condition,
  intensity = fg_quantity,
  plot = FALSE
)

qc_cvs(
  data = DIA_clean_uniprot,
  grouping = fg_id,
  condition = r_condition,
  intensity = fg_quantity,
  plot_style = "density",
  plot = TRUE
)

qc_cvs(
  data = DIA_clean_uniprot,
  grouping = fg_id,
  condition = r_condition,
  intensity = fg_quantity,
  plot_style = "violin",
  plot = TRUE
)

```

## Intensity distribution

Intensity distributions are plotted for the whole dataset.

```{r quality_control_log2_int_distr, fig.align= "center", message = FALSE, warning = FALSE}


qc_intensity_distribution(
  DIA_clean_uniprot,
  condrep,
  pep_grouping_key,
  intensity_log2,
  plot_style = "histogram"
)

```

## Missed cleavages


```{r quality_control_missed_cleav, fig.align= "center", message = FALSE, warning = FALSE}
qc_missed_cleavages(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_nr_of_missed_cleavages,
  fg_quantity,
  method = "intensity",
  plot = TRUE,
  interactive = FALSE
)

qc_missed_cleavages(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_nr_of_missed_cleavages,
  fg_quantity,
  method = "count",
  plot = TRUE,
  interactive = FALSE
)
```

## Peptide types

Peptide type (tryptic, semi-tryptic, non-tryptic) distributions are even throughout the different samples.

```{r quality_control_pep_type, fig.align = "center", message = FALSE, warning = FALSE}

qc_peptide_type(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_type,
  intensity = fg_quantity,
  method = "count",
  plot = TRUE,
  interactive = FALSE
)

qc_peptide_type(
  DIA_clean_uniprot,
  condrep,
  fg_id,
  pep_type,
  intensity = fg_quantity,
  method = "intensity",
  plot = TRUE,
  interactive = FALSE
)

```

## Number of peptide IDs per sample

The numbers of identified peptides are consistent throughout the different samples.

```{r quality_control_pep_ID, fig.align = "center", message = FALSE, warning = FALSE}

DIA_raw$condrep <- paste(DIA_raw$r_condition, DIA_raw$r_replicate, sep = "_")


qc_ids(DIA_raw, condrep, pep_grouping_key, condition = r_condition, intensity = fg_quantity)

```

## Principal component analysis (PCA)

```{r PCA, message = FALSE, warning = FALSE}

DIA_clean_uniprot %>%
qc_pca(condrep, 
       pep_grouping_key, 
       intensity_log2, 
       r_condition)

```



## corelation_map


```{r correlation_map, message = FALSE, warning = FALSE}

qc_sample_correlation(
  data = DIA_clean_uniprot,
  sample = r_file_name,
  grouping = fg_id,
  intensity_log2 = intensity_log2,
  condition = r_condition
)

```







#Data analysis
## Volcano plots peptide lvl


```{r Volcano_plot, message = FALSE, warning = FALSE}

  


Volcano_input <- DIA_clean_uniprot_summed %>%
  unique() %>%
  assign_missingness(r_file_name,
                     r_condition,
                     pep_stripped_sequence,
                     summed_intensity_log2,
                     ref_condition = "0_mM",
                     retain_columns = c(pg_protein_accessions, pep_stripped_sequence, start, end, pep_type, sequence, length))


Volcano_input = Volcano_input %>%
  dplyr::filter(is.na(sequence)== F)



#differential analysis

t_test_result_mod_t_test_pep <- calculate_diff_abundance(data = Volcano_input,
                                r_file_name,
                                r_condition,
                                pep_stripped_sequence,
                                summed_intensity_log2,
                                missingness,
                                comparison = comparison,
                                ref_condition = "0_mM",
                                method = "moderated_t-test",
                                retain_columns = c(pg_protein_accessions, pep_stripped_sequence, start, end, pep_type, sequence, length))



condition_vulcano <- c("0.004_mM_vs_0_mM")
t_test_result_mod_t_test_pep_comp <- t_test_result_mod_t_test_pep[t_test_result_mod_t_test_pep$comparison %in% condition_vulcano, ]


volcano_plot(t_test_result_mod_t_test_pep_comp, pep_stripped_sequence, diff, adj_pval, method = "target", target_column = pg_protein_accessions, target = c("P26365") ,  x_axis_label = "log2(fold change) Mercapto vs. DMSO", title = "Volcano plot peptide level", y_axis_label = "-log10(q-value)", log2FC_cutoff = 1, significance_cutoff = 0.05, interactive = T)



# Specify the output path
#output_path_diff_abundance <- "add/here.csv" 

# Save the dataframe as a CSV file
#write.csv(t_test_result_mod_t_test_pep, file = output_path_diff_abundance, row.names = FALSE)


```



##LiP-Quant


```{r LiP_Quant, message = FALSE, warning = FALSE}

DIA_clean_uniprot_summed$conc_frag <- as.numeric(gsub("_mM", "", DIA_clean_uniprot_summed$r_condition))




fit <- parallel_fit_drc_4p(
  DIA_clean_uniprot_summed,
  r_file_name,
  pep_stripped_sequence,
  summed_intensity_log2,
  conc_frag,
  filter = "post",
  n_replicate_completeness = 3,
  n_condition_completeness = 2,
  correlation_cutoff = 0.8,
  log_logarithmic = TRUE,
  retain_columns =  c(pg_protein_accessions, start, end, pep_type)
)


# Step 1: Group by protein accession
top_100_proteins <- DIA_clean_uniprot_summed %>%
  group_by(pg_protein_accessions) %>%             # Group by protein accession
  summarize(total_intensity = mean(pg_quantity)) %>%  # Summarize intensity for each protein
  top_n(100, wt = total_intensity) %>%                # Select the top 100 proteins based on intensity
  pull(pg_protein_accessions)                          # Get the list of top 100 protein accessions

# Step 2: Filter the original dataframe to retain only the top 100 proteins and all their peptides
df_top_100 <- DIA_clean_uniprot_summed %>%
  filter(pg_protein_accessions %in% top_100_proteins)




drc_fit <- fit_drc_4p(
  data = df_top_100,
  sample = r_file_name,
  grouping = pep_stripped_sequence,
  response = summed_intensity_log2,
  dose = conc_frag,
  n_replicate_completeness = 3,
  n_condition_completeness = 2,
  retain_columns = c(pg_protein_accessions, start, end, pep_type)
)


fit_copy <- fit


# Check for columns that contain lists
is_list_column <- sapply(fit_copy, is.list)


# Convert list columns to strings (for all columns that are lists)
fit_copy[is_list_column] <- lapply(fit_copy[is_list_column], function(col) {
  sapply(col, function(x) paste(x, collapse = ";"))
})




# Specify the output path
output_path_fit <- "/Volumes/biol_bc_picotti_1/Maximilian_D/028_Robot2_higher_conc/Frag_2/fit.csv" 

# Save the dataframe as a CSV file
write.csv(fit_copy, file = output_path_fit, row.names = FALSE)




fit_filtered <- fit[!is.na(fit$rank), ]





# Model plotting
drc_4p_plot(fit_filtered,
            grouping = pep_stripped_sequence,
            dose = conc_frag,
            response = summed_intensity_log2,
            targets = "all",
            unit = "%")




```

